Evolution of ORM Systems: Formal Foundations

### 2013-01-19 doc/description.tex
dropX with data

A lot of introduced transformations can be applied on an ordered set. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}
 \end{cases}\\
 y \lhd Software \vee y \lhd Application \vee y \lhd Database
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant.

@@

\paragraph{Rename Property and Association} Both renaming operations can change a name of the property, association respectively if there are no name collisions:
\begin{align*}
	\mathbf{renProp} : Label \times Label \times Label \times Application \rightarrow Application \\
	renProp(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n
	}{\begin{gathered}
		p.name = l_n 
	\end{gathered}}\\
	l_c, l_o, l_n \in Label, a \lhd Application
\end{align*}
\begin{align*}
	\mathbf{renAssoc} : Label \times Label \times Label \times Application \rightarrow Application \\
	renAssoc(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists b \in associations(c) : name(b) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n
	}{\begin{gathered}
		b.name = l_n 
	\end{gathered}}
	l_c, l_o, l_n \in Label, a \lhd Application
\end{align*}

@@

\subsubsection{Database decomposition}
\paragraph{Remove table} The function removes a table if there are no stored data. This is because the data lost is undesirable in case of evolution. The consistency of foreign keys references has to be verified too.
\begin{align*}
	\mathbf{dropTable} : Label \times Database \rightarrow Database \\
	dropTable(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \not \exists r \in data(d) : owningTable(r) = l \\  \not \exists j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) = l }{ tables(d) \setminus \rho(c)}
\end{align*}

@@

\paragraph{Remove Data and Column} The approach we declare to introduce in this paper was declared to be data and information preserving, nevertheless there are cases when data can be deleted from the database without data lost as they are e.g. already copied in another column as the understanding of data changed in the whole software. 
\begin{align*}
	\mathbf{RemoveColumAndData} : Label \times Label* \times Database \rightarrow Database \\
	RemoveColumnAndData(l_c, ps, d) = \begin{cases}
 	%\inference{ps = \varnothing}{d}\\
 	 %\inference{|ps| >1}{\begin{gathered} 	d \bullet RemoveColumnAndData(l_c, head(ps), d) \\ \bullet RemoveColumnAndData(l_c, tail(ps), d) \end{gathered}} \\ \\
 	\inference{|ps| = 1 \\ \wedge \exists t \in tables(d) \\ \wedge \exists c \in columns(t) : name(c) = l_c}{\begin{gathered}
	 	(\forall r, r \in tableData(d) \wedge reference(r) = l_c, \\ p \in pairs(r) \wedge name(p) = ps : pairs(r) \ ps) \\ \wedge columns(t) \ c  
\end{gathered}}
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, d \in Database
\end{align*}

@@

@@

\paragraph{Remove Property and its Data} The $RemovePropertyAndData$ transformation helps to remove a property together with data stored in the database. The evolution can cause data and information lost, to prevent this the transformation $RemovePropertyAndData$ should be used carefully.
\begin{align*}
	\mathbf{RemovePropertyAndData} : Label \times Label* \times State \rightarrow State \\
	\Phi(RemovePropertyAndData(L_c, ps), s) = application(s)\\ \circ remProperty(l_s, ps) \\
	\Psi(RemovePropertyAndData(L_c, ps), s) = \begin{cases}
		%\inference{ps = \varnothing}{d}\\
		%\inference{|ps| > 1}{\begin{gathered} database(s) \\ \bullet RemovePropertyAndData(L_c, head(ps)) \\ \bullet RemovePropertyAndData(L_c, tail(ps)) 	\end{gathered}}\\\\
 		\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet RemoveColumnAndData(l_c, ps)}\\\\
 	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) > 1}{database(s) \bullet RemoveTableAndData(l_c, ps)}
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, s \lhd Software
\end{align*}

>>>>

A lot of introduced transformations can be applied on an ordered set. The application proceeds on set consists in sequential application on set elements:
\begin{align*}
	f(x, y) = \begin{cases}
 		\inference{x = \varnothing}{y}\\
 		\inference{|x| > 1}{f(head(x), y) \cup f(tail(x), y}\\\\
 		\inference{|x| = 1}{effect}
 		\inference{otherwise}{Inconsistent y}
 \end{cases}\\
 y \lhd Software \vee y \lhd Application \vee y \lhd Database
\end{align*}
where $x$ is a set of parameters. The effect of a transformation is provided only in next sections as the repetition of sequential application will be redundant. The $otherwise$ branch is consider implicit for all introduced transformations and thus it is not part of transformation definitions too.

@@

\paragraph{Rename Property and Association} Both renaming operations can change a name of the property, association respectively if there are no name collisions:
\begin{align*}
	\mathbf{renProp} : Label \times Label \times Label \times Application \rightarrow Application \\
	renProp(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists p \in properties(c) : name(p) = l_o \\ \wedge \forall e \in properties(c) : name(e) \neq l_n
	}{\begin{gathered}
		p.name = l_n 
	\end{gathered}}\\
	l_c, l_o, l_n \in Label, a \lhd Application
\end{align*}
\begin{align*}
	\mathbf{renAssoc} : Label \times Label \times Label \times Application \rightarrow Application \\
	renAssoc(l_c, l_o, l_n, a) = \inference{\exists c \in classes(a) : name(c) = l_c \\ \wedge \exists b \in associations(c) : name(b) = l_o \\ \wedge \forall e \in associations(c) : name(e) \neq l_n
	}{\begin{gathered}
		b.name = l_n 
	\end{gathered}}
	\\ l_c, l_o, l_n \in Label, a \lhd Application
\end{align*}

@@

\subsubsection{Database decomposition}
\paragraph{dropValue}
\begin{align*}
	\mathbf{dropValue} : Label \times Row \times Database \rightarrow Database \\
	dropValue(l, r, d) = \inference{\exists p \in pairs(r) : reference(p) = l}{pairs(r) \setminus p}
	\\ l \in Label, r \lhd Row \wedge r \in data(d), d \lhd Database
\end{align*}

\begin{align*}
	\mathbf{dropRow} :  Row*  \times Database \rightarrow Database\\
	dropRow(r, d) = \inference{|r|=1}{data(d) \setminus r}
	\\ l \in Label, r \lhd Row \wedge r \in data(d), d \lhd Database
\end{align*}

\paragraph{Remove table} The function removes a table if there are no stored data. This is because the data lost is undesirable in case of evolution. The consistency of foreign keys references has to be verified too.
\begin{align*}
	\mathbf{dropTable} : Label \times Database \rightarrow Database \\
	dropTable(l,d) = \inference{ \exists t \in tables(d) : name(t) = l  \\ \wedge \forall r \in data(d) : owningTable(r) \neq l \\  \wedge (\forall j \in tables(d) \wedge f \in foreignKeys(j) : reference(f) \neq l) }{ tables(d) \setminus \rho(c)}
\end{align*}

\begin{align*}
	\mathbf{dropTableAndData} : Labe \times Database \rightarrow Database \\
	dropTableAndData(l, d) = \inference{\exists t \in tables(d) \\ \wedge ! \; associated(t, d)}{\begin{gathered}
	(\forall r \in tableData(t,d) : dropRow(r, d))\\ \bullet dropTable(t,d)
\end{gathered}}
	\\ l \in Label, d \lhd Database
\end{align*}

@@

@@

\paragraph{Remove Data and Column} The approach we declare to introduce in this paper was declared to be data and information preserving, nevertheless there are cases when data can be deleted from the database without data lost as they are e.g. already copied in another column as the understanding of data changed in the whole software. 
\begin{align*}
	\mathbf{DropColumnAndData} : Label \times Label* \times Database \rightarrow Database \\
	DropColumnAndData(l_c, ps, d) = \begin{cases}
 	%\inference{ps = \varnothing}{d}\\
 	 %\inference{|ps| >1}{\begin{gathered} 	d \bullet DropColumnAndData(l_c, head(ps), d) \\ \bullet DropColumnAndData(l_c, tail(ps), d) \end{gathered}} \\ \\
 	(\inference{|ps| = 1 \\ \wedge \exists t \in tables(d) \\ \wedge \exists c \in columns(t) : name(c) = l_c}{\begin{gathered}
	 	(\forall r, r \in tableData(d) \wedge reference(r) = l_c: \\d \bullet dropValue(ps, r)) \end{gathered}}) \\ \bullet dropColumn(l_c)  
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, d \in Database
\end{align*}

\begin{align*}
	\mathbf{DropKeyAndData} : Label \times Label* \times Database \rightarrow Database \\
	DropKeyAndData(l_f, ps, d) = \begin{cases}
 	%\inference{ps = \varnothing}{d}\\
 	 %\inference{|ps| >1}{\begin{gathered} 	d \bullet DropColumnAndData(l_c, head(ps), d) \\ \bullet DropColumnAndData(l_c, tail(ps), d) \end{gathered}} \\ \\
 	(\inference{|ps| = 1 \\ \wedge \exists t \in tables(d) \\ \wedge \exists f \in foreignKeys(t) : name(f) = l_f}{\begin{gathered}
	 	(\forall r, r \in tableData(d) \wedge reference(r) = l_f: \\d \bullet dropValue(ps, r)) \end{gathered}}) \\ \bullet dropKey(l_f)  
 \end{cases}
	\\
	l_f \in Label, ps \lhd Label*, d \in Database
\end{align*}

@@

\paragraph{Remove Property and its Data} The $RemovePropertyAndData$ transformation helps to remove a property together with data stored in the database. The evolution can cause data and information lost, to prevent this the transformation $RemovePropertyAndData$ should be used carefully.
\begin{align*}
	\mathbf{RemovePropertyAndData} : Label \times Label* \times State \rightarrow State \\
	\Phi(RemovePropertyAndData(L_c, ps), s) = application(s)\\ \circ remProperty(l_s, ps) \\
	\Psi(RemovePropertyAndData(L_c, ps), s) = \begin{cases}
		%\inference{ps = \varnothing}{d}\\
		%\inference{|ps| > 1}{\begin{gathered} database(s) \\ \bullet RemovePropertyAndData(L_c, head(ps)) \\ \bullet RemovePropertyAndData(L_c, tail(ps)) 	\end{gathered}}\\\\
 		\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) \leq 1}{database(s) \bullet DropColumnAndData(l_c, ps)}\\\\
 	\inference{\exists c \in classes(application(s)) : name(c) = l\\ 	\wedge \exists p \in properties(c) : name(p) = l \\ \wedge cardinality(p) > 1}{database(s) \bullet RemoveTableAndData(l_c, ps)}
 \end{cases}
	\\
	l_c \in Label, ps \lhd Label*, s \lhd Software
\end{align*}
